Reproducible development environments with rix

### 2024-10-01 paper.qmd
formatting
typos and update to latest version of packages

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clearly list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

@peng2011 introduced the idea of reproducibility being on a continuum: on one of
the ends of this continuum, we only have access to the paper describing the
studies, which is not reproducible at all. If in addition to this paper authors
make the original source code of the analysis available, reproducibility is
improved. Adding the original data improves reproducibility yet again. 
Finally, if to all this the authors add what Roger
Peng named the *linked and executable code and data*, we reach the gold standard
of reproducibility.

What is this *linked and executable code and data*? Another way to name this
crucial piece of the reproducibility puzzle is *computational environment*. The
computational environment is all the software required to actually run the
analysis. Here too, we can speak of a continuum. One could simply name and list
the software used: for example, the [R]{.proglang} programming language.
Sometimes, authors have the courtesy to also state the version of [R]{.proglang}
used. Some authors go further, and also list the packages used, and ideally with
their versions as well. Authors rarely state the operating system on which the
analysis was done, even though it has been shown that running the same analysis
with the same software but on different operating systems could lead to
different results, as described in @neupane2019. Authors also only very rarely
provide instructions to install the required tools and software in order to
reproduce their studies.

@@

There are tools available that authors can use to reach the gold stardard of a
fully reproducible study, or to successfully deploy the same development
environment on several machines. Let us first consider the task of listing the
software used. [R]{.proglang} provides the `sessionInfo()` function whose output
can be saved into a file. Below is an example output of `sessionInfo()`:

@@

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future), or co-authors, can read
this file and see which version of [R]{.proglang} was used, and which packages
(and their versions) were used as well. However, others would still need to
install the correct versions of each package manually. This can be in certain
cases challenging, especially if packages that depend on system-level
dependencies are required. For example, the [nloptr]{.pkg} package requires
either [cmake]{.proglang} on Linux or macOS, or a binary of [nlopt]{.proglang},
but this is not the case of Windows.

@@

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitely and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

@@

The advantage of using [Docker]{.pkg} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.pkg} images are actually almost complete Linux
operating systems, which means that system-level dependencies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
thus also made available to future reproducers of the study within the
[Docker]{.pkg} image. The image can also be shared.

@@

While [Docker]{.pkgs} offers a great solution to the reproducibility
problem, it can be awkward to work interactively from a running container:
containers can be changed at run-time, but any change will be lost once the
container is stopped. Also, while it is possible to launch graphical programs
from a container, this is generally not easy to set up and only works, as far as
we know, only on Linux operating systems. In most cases, users of
[Docker]{.pkg} will instead use the container to run a web server and the
web based application of a popular integrated development environment (IDE).
A common approach seems to be to
work on a study interactively using a standard installation of R and R packages,
and after the study is done, take some time to write the required `Dockerfile`
to provide an easy way for future replicators to reproduce the results of the
study. But this approach does not solve the problem of deploying the same
environment on several machines, and so ideally authors would develop using
a containerized development environment. Another challenge of using [Docker]{.pkg} is that,
as stated already, [Docker]{.pkg} images are actually almost complete
Linux operating systems, so experience with Linux is highly recommended if
one wishes to properly write a `Dockerfile`. Finally, one also needs to be careful
and write the `Dockerfile` in such a way that the build process will always result
in the same image. This is only possible if a specific digest of a base
image is used instead of a tagged version, but in practice, this is rarely done.

@@

@mcdermott2021 is an example of a scientific study that reached the gold
standard of reproducibility. The author of this paper set up an accompagnying
Github repository to the
paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
instructions to install the required software and then run the analysis. If we
take a closer look at this repository, we will notice that many of the tools
previously mentioned were used to capture the compatutational environment and
make it available to other researchers:

- Packages and their versions were listed and saved into an `renv.lock` file;
- A `Makefile` was used to run the whole analysis and compile the paper;
- A `Dockerfile` was used to provide the complete computational environment,
  including the right version of [R]{.proglang} and run the whole analysis
  easily.

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [Make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

@@

[Nix]{.pkg} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.pkg}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.pkg} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with
[Nix]{.pkg}, it suffices to declare that the [xlsx]{.pkg} package is needed
for the project, and [Nix]{.proglang} figures out automatically that
[Java]{.proglang} is required and installs and configures it. It all just
happens without any required intervention from the user. In this context, 
[xlsx]{.pkg} can be refered to as a component closure, and quoting @dolstra2004nix:

> The idea is to always deploy component closures: if we deploy a
> component, then we must also deploy its dependencies, their dependencies, and so on. That is, we must
> always deploy a set of components that is closed under
> the ''depends on'' relation. Since closures are selfcontained, they are the units of complete software
> deployment. After all, if a set of components is not
> closed, it is not safe to deploy, since using them might
> cause other components to be referenced that are missing on the target system.

But how does [Nix]{.pkg} achieve this? Where do these packages, or closures, come from? When installing a package using
[Nix]{.pkg}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
an environment for a project in a reproducible manner. The dependencies of these
derivations are also defined in other expressions, which will also get installed
if needed.

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
[Nix]{.pkg} package manager itself was designed with reproducibility in
mind. [Nix]{.pkg} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built (an important
side-note is in order here: while the functional aspect of [Nix]{.pkg} is quite
useful to ensure reproducibility, it can lead to situations where writing the
required [Nix]{.proglang} expression to build it can be much more difficult than usual: this is especially true for
packages that must download assets off the internet during installation, such as 
several BioConductor packages like [musData]{.pkg} for example. But this is only
a concern for the packagers, not users of the package that simply wish to install
and use the package). [Nix]{.pkg} also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
to be installed at the same time on the same system. Each variant of a package
has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
installed and use the correct one where adequate.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects. It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete, project-specific,
and reproducible environment and use it interactively, or non-interactively. It will
always be possible to rebuild this environmen in the future (as long as the `nixpkgs`
GitHub repository is online) to run an analysis sometime in the future.

@@

```
let
  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/976fa3369d722e76f37c77493d99829540d43845.tar.gz") {};
  system_packages = builtins.attrValues {
    inherit (pkgs) R ;
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
```

To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used (it should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository).
`system_pkgs` lists the packages that we wish to install and use with this
environment; in this example only the [R]{.proglang} programming language will
be installed, alongside all of its dependencies, and all of the dependencies’
dependencies. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).

@@

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.pkg} package manager for reproducibility, we wrote [rix]{.pkg}. [rix]{.pkg}
comes with the `rix()` function that helps generate such expressions quite
easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

Here, users do not need to look for the right commit of `nixpkgs`: it is enough
to specify the version of [R]{.proglang} they wish to use and the right
commit will be selected (a lookup table linking versions of R and `nixpkgs` commits
is included in the package). The versions of [R]{.proglang} packages that
were current at the time will get installed in that environment. But [rix]{.pkg} is quite flexible and
makes it easy to install specific versions of packages as well (which will get
installed from CRAN) as well as packages hosted on Github and
\LaTeX packages as well:

@@

The above call to `rix()` generates an expression that installs several
[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
older version of the package. Had the user not specified the version, the
version available as of the `nixpkgs` commit would be used), but also install
several TeXLive packages for authoring documents with \LaTeX, the development
versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
the RStudio integrated development environment. It should be noted that as of
writing, RStudio is not available on macOS through `nixpkgs`, so the above
expression will only work on Windows and Linux distributions. To make the
expression work on macOS, the "rstudio" string could be replaced by "code",
which would install VS Code instead, or "other", to use any other editor such as
Emacs or Vi. You can see the resulting `default.nix` file in the @sec-appendix.

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Essentially, each time one of the authors pushed to the master
branch, a Github Actions runner would be started, [Nix]{.proglang} would get
installed on said runner, and an environment containing the right packages was
generated from a `default.nix` file that is hosted on the same repository. The
paper would then get compiled using [Quarto]{.pkg} in this reproducible environment.

@@

The final feature that we provide is the ability to run one arbitrary function
from what we call a *subshell*. Suppose that you are running the latest version
of R, and that you wish or need to run one particular function from an older
release of R, or from an older package. Using the `with_nix()` function, it is
possible to run this single function inside the right envirnoment, and recover
its output in the main session. To illustrate, consider the following two lines
of code:

@@

We recommend `path_subshell` to be a subfolder of the main project. Using `rix()` we
generate a `default.nix` file that will provide a shell with R version 3.5.3
when built. We then use `rix_init()`, which generates an `.Rprofile` file in the
same subfolder as where the subshell's `default.nix` is located. This
`.Rprofile` file ensures that the R session that will be spawned from this
subshell does not get any interference from the main session it gets called
from:

@@

```{r, include = F}
rix_init(
   project_path = path_subshell,
   rprofile_action = "overwrite",
   message_type = "simple"
 )
```

>>>>

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clearly list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

@peng2011 introduced the idea of reproducibility being on a continuum: on one of
the ends of this continuum, we only have access to the paper describing the
studies, which is not reproducible at all. If in addition to this paper authors
make the original source code of the analysis available, reproducibility is
improved. Adding the original data improves reproducibility yet again. Finally,
if to all this the authors add what Roger Peng referred to as the *linked and
executable code and data*, we reach the gold standard of reproducibility.

*Linked and executable code and data* refer to the compiled source code scripts,
but for an interpreted language such as [R]{}{.proglang}, this would rather
refer to the *computational environment*. The computational environment is all
the software required to run the analysis. Here too, we can speak of a
continuum. One could simply name and list the software used: for example, the
[R]{.proglang} programming language. Sometimes, authors have the courtesy to
also state the version of [R]{.proglang} used. Some authors go further, and also
list the packages used, and ideally with their versions as well. Authors rarely
state the operating system on which the analysis was done, even though it has
been shown that running the same analysis with the same software but on
different operating systems could lead to different results, as described in
@neupane2019. Authors also only very rarely provide instructions to install the
required tools and software in order to reproduce their studies.

@@

There are tools available that authors can use to reach the gold standard of a
fully reproducible study, or to successfully deploy the same development
environment on several machines. Let us first consider the task of listing the
software used. [R]{.proglang} provides the `sessionInfo()` function whose output
can be saved into a file. Below is an example output of `sessionInfo()`:

@@

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future), or co-authors, can read
this file and see which version of [R]{.proglang} was used, and which packages
(and their versions) were used as well. However, others would still need to
install the correct versions of each package manually. This can be in certain
cases challenging, especially if packages that depend on system-level
dependencies are required. For example, the [nloptr]{.pkg} package requires a
binary of [nlopt]{.proglang} to be available, or if not, [cmake]{.proglang} on
Linux or macOS.

@@

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitly and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

@@

The advantage of using [Docker]{.pkg} is not only about having the right version
of [R]{.proglang} and [R]{.proglang} packages bundled alongside the analysis
scripts: [Docker]{.pkg} images are actually almost complete Linux operating
systems, which means that system-level dependencies also get bundled into the
image. This is especially useful for packages that rely on external system-level
dependencies to run, such as for [nloptr]{.pkg}, as these external system-level
dependencies are thus also made available to future replicators of the study
within the [Docker]{.pkg} image. The image can also be shared.

@@

While [Docker]{.pkgs} offers a great solution to the reproducibility problem, it
can be awkward to work interactively from a running container: containers can be
changed at run-time, but any change will be lost once the container is stopped.
Also, while it is possible to launch graphical programs from a container, this
is generally not easy to set up and only works, as far as we know, only on Linux
operating systems. In most cases, users of [Docker]{.pkg} will instead use the
container to run a web server and the web based application of a popular
integrated development environment (IDE).

A common approach seems to be to work on a study interactively using a standard
installation of R and R packages, and after the study is done, take some time to
write the required `Dockerfile` to provide an easy way for future replicators to
reproduce the results of the study. But this approach does not solve the problem
of deploying the same environment on several machines, and so ideally authors
would develop using a containerized development environment. Another challenge
of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
actually almost complete Linux operating systems, so experience with Linux is
highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
also needs to be careful and write the `Dockerfile` in such a way that the build
process will always result in the same image. This is only possible if a
specific digest of a base image is used instead of a tagged version, but in
practice, this is rarely done.

@@

@mcdermott2021 is an example of a scientific study that reached the gold
standard of reproducibility. The author of this paper set up an accompanying
Github repository to the
paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
instructions to install the required software and then run the analysis. If we
take a closer look at this repository, we will notice that many of the tools
previously mentioned were used to capture the compatutational environment and
make it available to other researchers:

- Packages and their versions were listed and saved into an `renv.lock` file;
- A `Makefile` was used to run the whole analysis and compile the paper;
- A `Dockerfile` was used to provide the complete computational environment,
  including the right version of [R]{.proglang}, to run the whole analysis
  easily.

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other software,
including the programming language that was used. Ideally, a build automation
tool such as [Make]{.pkg} should be added to the list. It should also be
noted that these requirements are not limited to the [R]{.proglang} programming
language. A very similar approach should be taken if one uses
[Python]{.proglang}, or any other programming language, for statistical analysis
instead. Furthermore, the previous discussion also implies that for an analysis
to actually be reproducible at all, it must be made using free and open source
software, as proprietary software cannot be freely shared with others that wish
to reproduce a study, and there's no guarantee that others will have access to
the proprietary software used.

@@

[Nix]{.pkg} is a package manager that can be used to install and build software
in a completely reproducible manner. As of writing, it contains more than
100.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
but also all the packages required for a project. The reason why one should use
[Nix]{.pkg} to install [R]{.proglang} packages, and not use the usual, built-in,
`install.packages()` [R]{.proglang} function instead, is that when installing a
package with [Nix]{.pkg}, [Nix]{.pkg} makes sure to install every dependency of
every package, whether this dependency is another [R]{.proglang} package or a
system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with [Nix]{.pkg}, it
suffices to declare that the [xlsx]{.pkg} package is needed for the project, and
[Nix]{.proglang} figures out automatically that [Java]{.proglang} is required
and installs and configures it. It all just happens without any required
intervention from the user. In this context, [xlsx]{.pkg} can be referred to as
a component closure, and quoting @dolstra2004nix:

> The idea is to always deploy component closures: if we deploy a component,
> then we must also deploy its dependencies, their dependencies, and so on. That
> is, we must always deploy a set of components that is closed under the
> ''depends on'' relation. Since closures are selfcontained, they are the units
> of complete software deployment. After all, if a set of components is not
> closed, it is not safe to deploy, since using them might cause other
> components to be referenced that are missing on the target system.

But how does [Nix]{.pkg} achieve this? Where do these packages, or closures,
come from? When installing a package using [Nix]{.pkg}, an expression written in
the [Nix]{.proglang} programming language gets downloaded from the `nixpkgs`
Github repository and evaluated. This expression contains a so-called
*derivation*. A derivation defines a build: its dependencies, commands to build
and install the package in question, and then an output. Most of the time, a
derivation downloads source code, builds the software from the source and then
outputs a compiled binary. Derivations are extremely flexible, and by writing
one's own, it is possible to define and build an environment for a project in a
reproducible manner. The dependencies of these derivations are also defined in
other expressions, which will also get installed if needed. This is why the 
[Nix]{.pkg} manager *knows* that [Java]{.proglang} is needed for [xlsx]{.pkg}.

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to use
a specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
a revision ensures that every package that Nix installs will always be at
exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

While pinning revisions is essential for reproducibility, this is not the only
reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
a functional package manager, meaning that it uses concepts from functional
programming to build software: no-side effects or global variables are allowed,
and the output is always the same if given the same inputs, regardless of where
or when a package gets built (an important side-note is in order here: while the
functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
can lead to situations where writing the required [Nix]{.proglang} expression to
build it can be much more difficult than usual: this is especially true for
packages that must download assets off the internet during installation, such as
several BioConductor packages like [musData]{.pkg} for example. But this is only
a concern for the packagers, not users of the package that simply wish to
install and use the package). [Nix]{.pkg} also allows for multiple versions (or
variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
system. Each variant of a package has a unique identifier, so that it is
possible to have two versions of [R]{.proglang} installed and use the correct
one where adequate. We refer to @dolstra2004nix for a more technical discussion
on [Nix]{.pkg}.

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects. It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete,
project-specific, and reproducible environment and use it interactively, or
non-interactively. It will always be possible to rebuild this environment in the
future (as long as the `nixpkgs` GitHub repository is online) to run an analysis
sometime in the future.

@@

```
let
  pkgs = import (fetchTarball 
    "https://github.com/NixOS/nixpkgs/archive/976fa336....tar.gz"
  ) {};
  system_packages = builtins.attrValues {
    inherit (pkgs) R ;
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
```

To define a variable in the [Nix]{.proglang} language, the `let` keyword is
used, so the first lines of this expression define the variables called `pkgs`
and `system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used (it should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository).
`system_pkgs` lists the packages that we wish to install and use with this
environment; in this example only the [R]{.proglang} programming language will
be installed, alongside all of its dependencies, and all of the dependencies’
dependencies. Finally, the `mkShell` function is used to build a development
shell containing the set of packages defined by the variable `system_packages`
and its `shellHook` is set to `R --vanilla`: this means that when entering the
shell, [R]{.proglang} will be started with the `--vanilla` flag (thus ignoring
any startup options).

@@

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.pkg} package manager for reproducibility, we developed [rix]{.pkg}.
[rix]{.pkg} comes with the `rix()` function that helps generate such expressions
quite easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

Here, users do not need to look for the right commit of `nixpkgs`: it is enough
to specify the version of [R]{.proglang} they wish to use and the right commit
will be selected (a lookup table linking versions of R and `nixpkgs` commits is
included in the package). The versions of [R]{.proglang} packages that were
current at the time will get installed in that environment. But [rix]{.pkg} is
quite flexible and makes it easy to install specific versions of packages (which
will get installed from CRAN) as well as packages hosted on Github and \LaTeX
packages:

@@

The above call to `rix()` generates an expression that installs several
[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
older version of the package. Had the user not specified the version, the
version available as of the `nixpkgs` commit would be used), but also installs
several TeXLive packages for authoring documents with \LaTeX, the development
versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
the RStudio integrated development environment. It should be noted that as of
writing, RStudio is not available on macOS through `nixpkgs`, so the above
expression will only work on Windows and Linux distributions. To make the
expression work on macOS, the "rstudio" string could be replaced by "code",
which would install VS Code instead, or "other", to use any other editor such as
Emacs or Vim. You can see the resulting `default.nix` file in the @sec-appendix.

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Each time one of the authors pushed to the master branch, a Github
Actions runner would be started, [Nix]{.proglang} would get installed on said
runner, and an environment containing the right packages was generated from a
`default.nix` file that is hosted on the same repository. The paper would then
get compiled using [Quarto]{.pkg} in this reproducible environment.

@@

The final feature that we provide is the ability to run one arbitrary function
from what we call a *subshell*. Suppose that you are running the latest version
of R, and that you wish or need to run one particular function from an older
release of R, or from an older package. Using the `with_nix()` function, it is
possible to run this single function inside the right environment, and recover
its output in the main session. To illustrate, consider the following two lines
of code:

@@

We recommend `path_subshell` to be a subfolder of the main project. Using
`rix()` we generate a `default.nix` file that will provide a shell with R
version 3.5.3 when built. When calling `rix()`, another function is also called,
`rix_init()`, which generates an `.Rprofile` file in the same subfolder as where
the subshell's `default.nix` is located. This `.Rprofile` file ensures that the
R session that will be spawned from this subshell does not get any interference
from the main session it gets called from.

@@
